---
title: "00_setup"
format: markdown_github
editor: source
---

# USAID GDW (R Track) - Session 00: Setup

## Learning objectives

-   Become familiar with RStudio's user interface
-   Setup snippets + script structure
-   Understand basic operations
-   Know how to install and load packages

## Quick note

For this and the remainder of our workshop sessions, we will be relying [Quarto docs](https://quarto.org) to facilitate the training. Quaro documents allow us to interact you to interact with the code you're writing while focusing on the documentation as well. Code is run within "chunks", which you can add through the shortcut CTRL + ALT + I (or through the GUI, Code \> Insert Chunk).

## Setup

For these sessions, we'll be using RStudio which is an IDE, "Integrated development environment" that makes it easier to work with R. For help on getting setup and installing packages, please reference [this guide](https://usaid-oha-si.github.io/corps/rbbs/2022/01/28/rbbs-0-setup.html) from the coRps R Building Blocks Series.

## Exploring RStudio IDE

RStudio may seem imposing at first, but in no time you'll know where everything is. Your RStudio will be broken down into 4 main components.

![20220214_rstudio_ide](https://github.com/USAID-OHA-SI/land_of_smiles/assets/8933069/4e4eef15-028d-4701-b058-3e000e5f4293)

-   **Source**: this box won't be here when you first open RStudio. If you go to File > New File > R Script, you will see this box. This is where you can write/save or open your R scripts.
-   **Console**: below Source, you will have your console. This is where all your R code is excuted and where the outputs are displayed. If you want to run a line as a one off, you can write it here. There is an tab here called Terminal, which is where you can run code through the command line, git for example.
-   **Environment**: moving to the upper right hand corner, you have your environment tab. This box contains all the datasets or objects you have stored in your current working session.
-   **Catch All**: this last box has a lot of different features. You can see a number of tabs, which allow you to expore your directories/files, this is where your graphs will print, and where the help files are located.


## Adding Snippets

Before we get too much further, we want to make sure you add some key Snippets to save you time and headache down the road. RStudio code snippets are predefined code shortcuts that can be used to quickly insert commonly used code blocks. The use of snippets can improve coding efficiency, reduce the time spent copying and pasting code from other scripts, and improve the readability of your code by providing a standardized format for your analytical scripts.

To create your own snippets in Rstudio, go to “Tools > Global Options > Code > Edit Snippets. This will open a file where you can define your custom snippets using a simple syntax. We recommend copying the snippets we've developed and adding them to your environment. You can copy our [SI snippets](https://gist.github.com/tessam30/fc775a2f917ea5d62de0f6724c4aeada) from here and paste them into the Snippet window. 


## Open a new script

We can start coding right in the Console, but the value of a coding language is being able to save and document workflows. Anything written in the Console will execute your code, but you can't easily save and share it. Instead, you want to open up a new R Script; in the IDE, go to File > New File > R Script or use the short cut CTRL + SHIFT + N. 

Each time you open a new script, we recommend organizing it, which will make it easier for you to think through and you/others to review in the future. We can organize our scripts with a similar flow to organizing a document, to clean up the script visually and help you and others better understand what each part of the script is used for. We can utilize comments to add headers into our script for organization.

While there is no one right way to organize a script, here is an example of how we tend to set up our scripts within OHA/SI.  

**Introduction**: includes the header matter including name of author, purpose of script, date, reference id, and any notes
**Load Dependencies and Libraries**: this is where you will identify which package to load
**Define global options or functions**: set up global variables as well (variables you want to call on throughout the script). such as the file location or today's date.
**Import Data**: import all the dataset you need for your script in this section, rather than doing it in piecemeal through our your script
**Data Cleaning**: where most of the munging occurs to get your data in the best place to analyze/visualize
**Analysis/Visualization**: where the magic happens!
**Export**: export any output datasets or visuals

Here is what the structure might look like in your script.

```
# PROJECT:  QUEST FOR THE HOLY GRAIL
# AUTHOR:   A.Chafetz | USAID
# PURPOSE:  figure out where to start our journey
# REF ID:   29a4991d 
# LICENSE:  MIT
# DATE:     2023-08-28
# UPDATED: 

# DEPENDENCIES ------------------------------------------------------------
  
  library(tidyverse)
  library(gagglr)
  library(glue)
  library(scales)
  library(extrafont)
  library(tidytext)
  library(patchwork)
  library(ggtext)
  

# GLOBAL VARIABLES --------------------------------------------------------
  
  ref_id <- "29a4991d" #id for adorning to plots, making it easier to find on GH
  
  get_metadata() #list of MSD metadata elements

# IMPORT ------------------------------------------------------------------
  
  df_msd <- si_path() %>%
    return_latest("OU_IM") %>% 
    read_psd()   
  

# MUNGE -------------------------------------------------------------------

  ...
  

# VIZUALIZE ---------------------------------------------------------------

  ...
  
# EXPORT ------------------------------------------------------------------

  ...
  

```

This is where the Snippets you setup earlier come in handy. You can use the shortcut from the snippet to add this structure to your script. If you start typing `rscript` you should see an autocomplete for the snippet and you can hit TAB; at that point, the structure will populate in your script.


## Basic Operations
You can perform basic operations in R, like addition and multiplication. To run code from the console, you will hit CTRL + ENTER. This will run the current line your cursor is on. To run multiple lines, you will highlight those lines and then hit CTRL + ENTER. You can also run the entire script by clicking the "Source" button on the top right corner, or run highlighted lines by clicking "Run."

```{r}
3*3
```

R is more powerful than a simple calculator as you can store values or results as objects (or variables) when we use the assignment operator, `<-`.

```{r}
#value of variable m is 3 x 3
m <- 3*3

#the city we want to reference
city <- "Cedar Rapids"

#a list of important partners
partners <- c("Yard Goats", "Loons", "Isotopes")

```

Its important to note that when ever we work with strings/character, we need to use quotation marks; otherwise, R will try to interpret the text that as an object.

In the code above, we've also introduced comments, which are notes we write for ourselves and others. R knows something is a comment (and not to evaluate it) by looking for a `#`; everything after a `#` is not run. We have also introduced vectors in the last object. If we are combining multiple values or even objects together, we use `c()` and list them out, separating each different piece with a comma.

On more importnat operator to introduce you to is "the pipe." The concept may be a bit tricky to start with, but it makes our life much easier. Base R introduce the pipe in version 4.0.0, "`|>`" and it has been around in the `tidyverse` for quite a while, "`%>%`", showing up first in the `magrittr` package.

In the example below, we want to make combine all the partners together into a single value with a comma separating them. We can write this out in a nested code, much like we would see in Excel. 

```{r}
#nested statement without pipes
paste0(toupper(partners), collapse = ", ")
```

Alternatively we could write this out so each statement gets added to a intermediate objects/steps.

```{r}
p2 <- toupper(partners) 

paste0(p2, collapse = ", ")

```

The pipe simplifies our life. We can use the pipe to pass object from one evaluation to the next, making our code more legible and easier to follow. We can read the code below as `partners` are converted to upper case and then pasted together. 

```{r}
#code with pipes, everything is evaluated stepwise
partners |> toupper() |> paste0(collapse = ", ")
```

All three of the options above give you the same results, but the piping simplies your code.

In addition to values and strings being stores as objects, we can also store data (as data frames) in objects

```{r}
(df <- data.frame(x = c("A", "B", "C"),
                 y = c(1, 2, 3)
))

(df$z <- df$y + m)

```

## Loading packages

Proprietary programs like Excel, Tableau, or Stata, build in and maintain all of the functions into the software. With open-source packages, on the other hand, we need to load different libraries or packages that are written by other organization or individual users. This structure allows different users to provide specific "solutions" to a problem. For example, if you are creating maps, you might want to load up the `sf` package, whereas you may not need that if you are only working with basic data munging, e.g.`dplyr`, `tidyr`, `stringr`. 

A package is a collection of functions. If we wanted to run a time series regression or do text analysis, we would need to install and load different packages created by other users. There are lots and lots of packages out there and we will keep largely to those that are in the ["tidyverse"](https://www.tidyverse.org/), a set of packages "share an underlying design philosophy, grammar, and data structures" maintained mostly by "super developers" at RStudio. The other set of packages we'll use and promote are those [we have build in OHA](https://usaid-oha-si.github.io/tools/) and solve issues we routinely face with workflow and data. It's important to note that every time you start up a new session, you will have to load the packages you need.

We have just been working with some base functions with in R, but we typically need to use external packages. To install those packages from CRAN, we can run `install.packages()` with the package specified in quotes. Let's install `dplyr`. If you already have a package install, rerunning this will update the package. You only need to install a package once, but you will need to load the package every time you start a new R session.

```{r}
#install the dplyr package
install.packages("dplyr")

#load the package into the environment to use it
library(dplyr) 

```


## Additional Resources

-   [coRps R Building Blocks Series - Setup Part 0](https://usaid-oha-si.github.io/corps/rbbs/2022/01/28/rbbs-0-setup.html)
-   [coRps R Building Blocks Series - Visualization Part 1](https://usaid-oha-si.github.io/corps/rbbs/2022/02/14/rbbs-2-visualization-part1.html)
-   [R for Data Science (Wickham + Grolemund) - Chapter 4 - Workflow: basics](https://r4ds.had.co.nz/workflow-basics.html)
